On the Limitations of Weighted Model Counting

### 2020-03-12 paper.tex
Feedback from Vaishak (not that I asked for it...)

%\section{Introduction}

@@

% TODO: we don't have to treat 0 as an exception we state that \/ of an empty set is 0 (and /\ of an empty set is 1).
% TODO: use bigvee and bigwedge directly with sets.
\begin{theorem} \label{thm:representation}
  Let $\mathbf{B}$ be a finite Boolean algebra. Then every $x \in \mathbf{B}
  \setminus \{ 0 \}$ can be uniquely expressed as
  \[
  x = \bigvee_{\text{atoms } a \le x} a.
  \]
\end{theorem}
\begin{proof}
  A simple consequence of the theorem that every finite Boolean algebra is
  isomorphic to a field of subsets of a set, where the cardinality of the set is
  equal to the number of atoms in the Boolean algebra.
\end{proof}

@@

@@

\todo[inline]{Allow weight to be zero}
\begin{definition} \label{def:wmc}
  Let $\mathbf{B}$ be a finite Boolean algebra, and let $M \subseteq \mathbf{B}$
  be its set of atoms. Let $L \subseteq \mathbf{B}$ be such that every atom $m
  \in M$ can be uniquely expressed as $m = \bigwedge_{i \in I} l_i$ for some $\{
  l_i \}_{i \in I} \subseteq L$, and let $w : L \to \mathbb{R}_{>0}$ be
  arbitrary. The \emph{weighted model count} $\WMC : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}

@@

@@

@@

\begin{theorem} \label{thm:extension} % TODO: cite the fact about atoms
  Let $\mathbf{B}$ be a finite Boolean algebra freely generated by some set of
  `literals' $L$, and let $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$ be an
  arbitrary measure. We know that $\mathbf{B}$ has $n = 2^{|L|}$ atoms. Let
  $(a_i)_{i=1}^n$ be those atoms. Let $L' = L \cup \{ \phi_i \mid i \in [n] \}$
  be the set $L$ extended with $n$ new literals. Let $\mathbf{B'} =
  \mathbf{C}/I$ be the quotient of the Boolean algebra $\mathbf{C}$ freely
  generated by $L'$ and the ideal $I$ generated by
  \[
    \{\phi_i \land \phi_j \mid i, j \in [n], i \ne j \} \cup \{ a_i \land \phi_j
    \mid i, j \in [n], i \ne j \}.
  \]
  Let $h : \mathbf{B} \to \mathbf{B'}$ be the Boolean homomorphism defined by
  $h(l) = l$ for all $l \in L$ (note that this indeed defines a homomorphism).
  Let $w : L' \to \mathbb{R}_{> 0}$ be defined by
  \[
    w(l) = \begin{cases}
      m(a_i) & \text{if } l = \phi_i \text{ for some } i \in [n] \\
      1 & \text{otherwise}
    \end{cases}
  \]
  for all $l \in L'$. This defines a WMC measure $m' : \mathbf{B'} \to
  \mathbb{R}_{\ge 0}$. Furthermore,
  \[
    m(a) = m'(h(a)).
  \]
\end{theorem}

In simpler terms, any measure can be computed using WMC by extending the Boolean
algebra with more literals and connecting the two via a homomorphism. More
precisely, we are given the red part in
\[
  \begin{tikzcd}
    & \textcolor{red}{\mathbb{R}_{\ge 0}} & & \\
    \textcolor{red}{\mathbf{B}} \arrow[red]{ur}{m} \arrow{rr}{h} & & \mathbf{B'}
    \arrow{ul}[swap]{m'} & \\
    \textcolor{red}{L} \arrow[Subset,red]{u}{} \arrow[Subset]{rr}{} & & L'
    \arrow[Subset]{u}{} \arrow{r}{w} & \mathbb{R}_{>0}
  \end{tikzcd}
\]
and construct the black part in such a way that the triangle commutes.

@@

% TODO: proof that n is the smallest number (in the worst case, without
% independence assumptions)
%\begin{proposition}
%  In \cref{thm:extension}, if $L'$ has less than $n$ additional elements, then
%  there exists a measure $$
%\end{proposition}

% TODO: describe how my approach is better in the worst case than that by
% Darwiche

>>>>

\section{Introduction}

% Feedback: What are the main claims, what are the main takeaways, intuitive
% [???] of theorems to follow. To do this, we appeal to algebraic constructions
% to define the main concepts for introducing measures on Boolean algebras.

@@

% TODO: we don't have to treat 0 as an exception we state that \/ of an empty
% set is 0 (and /\ of an empty set is 1).
% TODO: use bigvee and bigwedge directly with sets.
\begin{theorem} \label{thm:representation}
  Let $\mathbf{B}$ be a finite Boolean algebra. Then every $x \in \mathbf{B}
  \setminus \{ 0 \}$ can be uniquely expressed as
  \[
  x = \bigvee_{\text{atoms } a \le x} a.
  \]
\end{theorem}
\begin{proof}
  A simple consequence of the theorem that every finite Boolean algebra is
  isomorphic to a field of subsets of a set, where the cardinality of the set is
  equal to the number of atoms in the Boolean algebra.
\end{proof}

@@

% TODO: Feedback: Probably a good idea to explain how elements of the Boolean
% algebra map to models in WMC formulations. Probably do an example:
% 0.3 p; 0.2 q => M = <p, qbar> has weight o3 * (1-0.2)
% then it corresponds to an atom in an algebra defined over the language defined
% over propositions p and q.

@@

\todo[inline]{Allow weight to be zero}
\begin{definition} \label{def:wmc}
  Let $\mathbf{B}$ be a finite Boolean algebra, and let $M \subseteq \mathbf{B}$
  be its set of atoms. Let $L \subseteq \mathbf{B}$ be such that every atom $m
  \in M$ can be uniquely expressed as $m = \bigwedge_{i \in I} l_i$ for some $\{
  l_i \}_{i \in I} \subseteq L$, and let $w : L \to \mathbb{R}_{>0}$ be
  arbitrary. The \emph{weighted model count} $\WMC : \mathbf{B} \to
  \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}
% TODO: Feedback: maybe NWMC or WMC^*, as nWMC looks like a typo. Also clarify
% the relationship between NWMC and the probability induced by a WMC
% formulation, i.e., Pr_{\Delta, w}(q) = \frac{WMC(\Delta \land q,
% w)}{WMC(\Delta, w)}.

@@

% TODO: Feedback: Can you say something here about factorized vs non-factorized
% weight function definitions? That is, factorized is when w maps literals to
% R_>=0, non-factorized is when w maps models to R_>=0 and show:
% a) come up with nice example when non-factorized weights are intuitive
% b) what if weight functions are negative/complex?
% c) clarify that the factorized definition have is w.r.t. models, in case some
% one gets confused [It doesn't have to be, if the BA is not free -- P.]
% d) can you say something about WMI

@@

% TODO: Feedback: My general remark here is that this section looks very [???]
% for SRL/WMC/PRM readers. The first-order [???] may not be justified if we are
% only doing essentially propositional models. If we also consider full FoL,
% i.e., infinite domains, and make this a separate paper for KR-type audience.
% That is, let us do the WMC/WMI bit separately and do FO-version in another
% paper for KR track. Title for second paper: "Measures on first-order
% structures via WMC" (or similar...).

@@

\begin{theorem} \label{thm:extension} % TODO: cite the fact about atoms
  Let $\mathbf{B}$ be a finite Boolean algebra freely generated by some set of
  `literals' $L$, and let $m : \mathbf{B} \to \mathbb{R}_{\ge 0}$ be an
  arbitrary measure. We know that $\mathbf{B}$ has $n = 2^{|L|}$ atoms. Let
  $(a_i)_{i=1}^n$ denote those atoms in some arbitrary order. Let $L' = L \cup
  \{ \phi_i \mid i \in [n] \}$ be the set $L$ extended with $n$ new literals.
  Let $\mathbf{B'}$ be the unique Boolean algebra with
  \[
    \{ a_i \land \phi_i \mid i \in [n] \} \cup \{ a_i \land \neg \phi_i \mid i
    \in [n] \}
  \]
  as its set of atoms. Let $\iota : \mathbf{B} \to \mathbf{B'}$ be the inclusion
  homomorphism (i.e., $\iota(a) = a$ for all $a \in \mathbf{B}$). Let $w : L'
  \to \mathbb{R}_{> 0}$ be defined by
  \[
    w(l) = \begin{cases}
      \frac{m(a_i)}{2} & \text{if } l = \phi_i \text{ or } l = \neg\phi_i \text{
        for some } i \in [n] \\
      1 & \text{otherwise}
    \end{cases}
  \]
  for all $l \in L'$, and note that this defines a WMC measure $m' : \mathbf{B'}
  \to \mathbb{R}_{\ge 0}$. Then
  \[
    m(a) = (m' \circ \iota)(a)
  \]
  for all $a \in \mathbf{B}$.
\end{theorem}

In simpler terms, any measure can be computed using WMC by extending the Boolean
algebra with more literals. More precisely, we are given the red part in
\[
  \begin{tikzcd}
    \textcolor{red}{\mathbb{R}_{\ge 0}} & & \\
    \textcolor{red}{\mathbf{B}} \arrow[red]{u}{m} \arrow{r}{\iota} &
    \mathbf{B'} \arrow{lu}[swap]{m'} & \\
    \textcolor{red}{L} \arrow[Subset,red]{u}{} \arrow[Subset]{r}{} & L'
    \arrow[Subset]{u}{} \arrow{r}{w} & \mathbb{R}_{>0}
  \end{tikzcd}
\]
and construct the black part in such a way that the triangle commutes.

@@

% TODO: proof that n is the smallest number (in the worst case, without
% independence assumptions)
\begin{proposition}
  In the worst case (i.e., with no independence assumptions), adding $2^{|L|}$
  literals is the best we can do.
%  In \cref{thm:extension}, if $L'$ has less than $n$ additional elements, then
%  there exists a measure $$
\end{proposition}

% TODO: describe how my approach is better in the worst case than that by
% Darwiche
% TODO: what about ENC2?
Let us note how our lower bound on the number of added literals compares to ENC1
\cite{DBLP:conf/kr/Darwiche02} and ENC2
